---
title: "Implementing the BYM2 Model in Stan"
format:
  html:
    code-copy: true
    code-overflow: wrap
    toc: true
editor_options: 
  chunk_output_type: console
---

### The BYM2 Model: Riebler et al, 2016

The ICAR model assumes complete spatial correlation between regions.
This simplifying assumption makes it cheap to compute;  the CAR model,
which adds a weight to each neighbor pair to account for the strength of
the spatial correlation, requires computing the inverse of the
adjacency matrix, an cubic operation.  For maps with more than a handful of regions,
this is prohibitively expensive, computationally.   The Besag York Mollié (BYM)
model was developed to address this problem; the BYM2 model builds on this model
and subsequent refinements.

* Poisson regression + component for extra variation which
combines spatial and non-spatial components $\phi + \theta$ as
$$\left( (\sqrt{\, {\rho} / s}\, \ )\,\phi^* + (\sqrt{1-\rho})\,\theta^* \right) \sigma $$
where:
  + $\sigma\geq 0$ is the overall standard deviation.
  + $\rho \in [0,1]$ - proportion of spatial variance.
  + $\phi^*$ is the ICAR component.
  + $\theta^* \sim N(0, 1)$ is the vector of ordinary random effects
  + $s$ is a scaling factor s.t. $\operatorname{Var}(\phi_i) \approx 1$; $s$ is data.

* Follows previous BYM modifications, notably Leroux 2000,
which has a single scale parameter plus mixing parameter.

* Adds a scaling factor on the ICAR component so that for each element $i$ in vectors $\theta$ and $\phi$,
$\operatorname{Var}(\theta_i) \approx \operatorname{Var}(\phi_i) \approx 1$.  The scaling factor is derived from the adjacency matrix (neighbors graph).

## Notebook Setup

### Libraries and helper functions.

```{r}
library(ggplot2)
library(igraph)
library(Matrix)
library(parallel)
library(sf)
library(spdep) |> suppressPackageStartupMessages()
library(tidyverse) |> suppressPackageStartupMessages()

library(cmdstanr)
library(posterior)
library(bayesplot)
library(loo)
cores = floor(parallel::detectCores() / 2)
source("utils_bym2.R")
source("utils_dataviz.R")
options(width = 120)
```

### Load NYC Dataset

```{r}
nyc_geodata = st_read(file.path("..", "data", "nyc_study.geojson"))
names(nyc_geodata)
```

## New York City Neighbor Graph

New York city consists of several islands; only the Bronx is part of the mainland; Brooklyn and Queens are part of Long Island, plus smaller islands City Island, Roosevelt Island, and the Rockaways.

*This is a problem for the ICAR model, which operates on a fully connected graph (single component)*

```{r}
nyc_nbs = poly2nb(nyc_geodata)
summary(nyc_nbs)
table(n.comp.nb(nyc_nbs)$comp.id)
```

* For the NYC analysis paper, we hand edited the map of NYC (in R) to create a fully connected network graph.

* For this notebook we restrict our attention to Brooklyn, the largest borough in NYC, which is a single network component.

```{r}
brooklyn_geodata = nyc_geodata[nyc_geodata$BoroName=='Brooklyn', ]
brooklyn_nbs = poly2nb(brooklyn_geodata)
brooklyn_coords = st_coordinates(st_centroid(brooklyn_geodata['geometry']))
plot(st_geometry(brooklyn_geodata), col='skyblue')
plot(brooklyn_nbs, coords=brooklyn_coords, add=TRUE, pch=20, cex=0.6)
```

```{r}
summary(brooklyn_nbs)
```

## From ICAR to BYM2

* The BYM2 model combines both ICAR component $\phi$ and ordinary random effects $\theta$ as
$$\left( (\sqrt{\, {\rho} / s}\, \ )\,\phi^* + (\sqrt{1-\rho})\,\theta^* \right) \sigma $$

* Parameter $\rho$ answers the question:  how much of the observed variance is spatial?

* Don't need to run analysis, e.g. Moran's I - the model sorts it out for you.

## BYM2 Model:  `bym2.stan`

This file is in directory `stan/bym2.stan`.

```{r}
bym2_model_file = file.path("..", 'stan', 'bym2.stan')
cat(readLines(bym2_model_file), sep="\n")
```

## Preparing the data inputs

### Get edgeset

- Compute this automatically from `nyc_geodata` spatial geometry component
  + Python package `libpysal`
  + R package `spdep`

```{r}
brooklyn_nbs_adj = nbs_to_adjlist(brooklyn_nbs)
brooklyn_nbs_adj[1:2, 1:10]
```

### Compute scaling factor `tau`

The BYM2 model component gamma combines an vector of non-spatial random effects `theta`,
and the vector of spatial random effects `phi`.
The prior on `theta` is `std_normal()` thus the its variance is approximately $1$.
The prior on `phi` is the ICAR component, the spatial covariance matrix.
To make the variance of the `phi` be approximately $1$,
Riebler et. al. recommend scaling it by the geometric mean of
the variance of the spatial adjacency matrix.
This quantity is `tau`, and is computed by helper function `get_scaling_factor`.

```{r}
tau = get_scaling_factor(brooklyn_nbs)
tau
```

#### Assemble the input data 

```{r}
design_mat <- as.data.frame(brooklyn_geodata) %>%
  select(pct_pubtransit, med_hh_inc, traffic, frag_index) %>%
  mutate(med_hh_inc = log(med_hh_inc),
         traffic = log(traffic))
summary(design_mat)
```

```{r}
bym2_data <- list(
  N = nrow(brooklyn_geodata),
  y = as.integer(brooklyn_geodata$count),
  E = as.integer(brooklyn_geodata$kid_pop),
  K = 4,
  xs = design_mat,
  N_edges = ncol(brooklyn_nbs_adj),
  neighbors = brooklyn_nbs_adj,
  tau = tau
)
```

## Workflow:  Prior Predictive Checking

Prior and posterior predictive checks are two cases of the general concept of predictive checks,
just conditioning on different things (no data and the observed data, respectively).
Posteior predictive checks were introduced in the Stan model building workflow notebook,
which shows how to generate a new dataset (`y_rep`) in the generated quantities block
based on the sample estimate of the model parameters.
Comparison of the replicated datasets with the observed datasets
shows how well or poorly the model captures the essential
characteristics of the data.

[Prior predictive checks](https://mc-stan.org/docs/stan-users-guide/posterior-predictive-checks.html#prior-predictive-checks)
simulate data directly from the prior, in the absense of any observed data.
The resulting datasets, often called `y_sim`, instead of `y_rep`,
show the possible range of data that is consistent with the priors.
This can be used to diagnose priors which are too strong or too weak.

The Stan User's Guide shows
[how to code](https://mc-stan.org/docs/stan-users-guide/posterior-predictive-checks.html#prior-predictive-checks)
in order to check all priors.
Unlike posterior predictive checks, in order to efficiently do the simulation,
it is necessary to rewrite the model.
In this notebook, we only want to verify that the BYM2 model does in fact properly
scale the variance of `phi`, so we only need to simulate the 
convolved spatial and random effects component `gamma`,
which will allow us to check `theta`, `phi`, and `rho`.


### Checking effect of scaling `phi` by `tau`.


To examine the effect of the scaling factor, we simulate
the parameters `gamma`, `phi`, `theta`, and `rho`
using a simplified version of the model that excludes
both the observed data (`y` and `xs`) and the Poisson regression component.
This allows us to assess the resulting priors on `phi`, `theta`, and `rho`
*before incorporating any data*.

```{r}
bym2_gamma_ppc_file = file.path("..", 'stan', 'bym2_gamma_ppc.stan')
cat(readLines(bym2_gamma_ppc_file), sep="\n")
```

#### Model compilation

```{r}
bym2_gamma_ppc_mod = cmdstan_model(stan_file=bym2_gamma_ppc_file)
```

#### Run model, check variances

In the BYM2 model, phi is scaled by the square root of $1$/`tau`
since the ICAR component `phi` follows a multivariate normal distribution
with covariance matrix $\Sigma$, whereas `theta` is a vector of standard normal variables.

```{r}
bym2_gamma_ppc_fit = bym2_gamma_ppc_mod$sample(data=bym2_data, parallel_chains=cores, refresh=0)
```

```{r}
phi <- as_draws_matrix(bym2_gamma_ppc_fit$draws("phi"))
phi_var_by_region <- apply(sqrt(1/tau) * phi, 2, var)  # variance across draws for each region
print(paste("mean variance of phi scaled:", mean(phi_var_by_region)))
theta <- as_draws_matrix(bym2_gamma_ppc_fit$draws("theta"))
theta_var_by_region <- apply(theta, 2, var)  # variance across draws for each region
print(paste("mean variance of theta:", mean(theta_var_by_region)))
```

#### Check `rho`

The fitted model also provides an estimate of `rho`, which has prior

```stan
rho ~ beta(0.5, 0.5);
```

As expected, before seeing any data, the value of `rho` matches the prior.

```{r}
rho <- as_draws_array(bym2_gamma_ppc_fit$draws("rho"))
mean(rho)
```


### Comparison with the Leroux model

When the scaling factor `tau` is 1.0,  the BYM2 model is equivalent to the Leroux model, and in this case, we can compute the variance of phi directly.

```{r}
print(paste("mean variance of phi:", mean(apply(phi, 2, var))))
```

## Fitting the BYM2 Model on the Brooklyn data

#### Model is compiled (as needed) on instantiation 

```{r}
bym2_mod = cmdstan_model(stan_file=bym2_model_file)
```

#### Run the NUTS-HMC sampler, summarize results

```{r}
bym2_fit = bym2_mod$sample(data=bym2_data, parallel_chains=cores, refresh=0)
bym2_fit$print(variables = c('beta_intercept', 'beta0', 'betas[1]', 'betas[2]', 'betas[3]', 'betas[4]', 'sigma', 'rho')) 
```

## Model Checking

### Visualizing the ICAR component

To see the spatial structure of the data, we plot the covariance matrix of `phi`.
We expect neighboring regions to be correlated, and distant regions to be uncorrelated.
To do this, we have written a helper function, `plot_icar_corr_matrix`,
which plots just the upper half of the correlation matrix of `phi`,
with red indicating positive correlation and blue indicating negative correlation.
The census tracts are listed roughly in geographic order, from north-to-south and
east-to-west, thus adjacent elements along the diagonal are spatially adjacent.

```{r}
phi_bym2 <- as_draws_matrix(bym2_fit$draws("phi"))
plot_icar_corr_matrix(phi_bym2, 'BYM2 spatial correlation')
```

##### Comparison to simple ICAR model 

```{r}
pois_icar_mod = cmdstan_model(stan_file=file.path("..", 'stan', 'poisson_icar.stan')) 
pois_icar_fit = pois_icar_mod$sample(data=bym2_data, parallel_chains=cores, refresh=0)
phi_icar <- as_draws_matrix(pois_icar_fit$draws("phi"))
plot_icar_corr_matrix(phi_icar, 'ICAR spatial correlation')
```

Compared to the covariance matrix of `phi` for the BYM2 model,
the spatial correlations are less sharp, i.e. the addition of
component `theta` allows `phi` to more strongly identify
the spatial correlation structure of the data.

## Model Comparison: BYM2 vs Poisson + ICAR vs. Poisson + RE

#### Poisson + ICAR model

```{r}
pois_icar_fit$print(variables = c('beta_intercept', 'beta0', 'betas[1]', 'betas[2]', 'betas[3]', 'betas[4]', 'sigma'))
```

#### Poisson + RE model

```{r}
pois_re_mod = cmdstan_model(stan_file=file.path("..",  'stan', 'poisson_re.stan')) 
pois_re_fit = pois_re_mod$sample(data=bym2_data, parallel_chains=cores, refresh=0)
pois_re_fit$print(variables = c('beta_intercept', 'beta0', 'betas[1]', 'betas[2]', 'betas[3]', 'betas[4]', 'sigma'))
```

### Compare overall variance

In the these models `sigma` determines the overall magnitude of variation
in the combined spatial (`phi`) and non-spatial(`theta`) random effects.
The BYM2 model adds parameter `rho`, the proportion of variance which is spatially structured.

```{r}
cat("BYM2")
bym2_fit$print(variables = c('sigma', 'rho'))

cat("Poisson ICAR")
pois_icar_fit$print(variables = c('sigma'))

cat("Poisson Ordinary Random Effects")
pois_re_fit$print(variables = c('sigma'))
```

### Compare Posterior predictive check coverage

```{r}
y_rep_bym2 <- as_draws_matrix(bym2_fit$draws("y_rep"))
cat("PPC BYM2")
ppc_central_interval(y_rep_bym2, bym2_data$y)

y_rep_icar <- as_draws_matrix(pois_icar_fit$draws("y_rep"))
cat("PPC Poisson + ICAR")
ppc_central_interval(y_rep_icar, bym2_data$y)

y_rep_re <- as_draws_matrix(pois_re_fit$draws("y_rep"))
cat("PPC Poisson + RE")
ppc_central_interval(y_rep_re, bym2_data$y)
```

### PPC plots

#### BYM2 model

```{r}
ppc_y_yrep_overlay(y_rep_bym2, bym2_data$y,
                     'BYM2 model PPC\ny (blue dot) vs. y_rep (orange 50% central interval, grey full extent)')
```


#### ICAR model

```{r}
ppc_y_yrep_overlay(y_rep_icar, bym2_data$y,
                     'Poisson + ICAR model PPC\ny (blue dot) vs. y_rep (orange 50% central interval, grey full extent)')
```

#### RE model

```{r}
ppc_y_yrep_overlay(y_rep_re, bym2_data$y,
                     'Poisson + RE model PPC\ny (blue dot) vs. y_rep (orange 50% central interval, grey full extent)')
```

## Discussion

The BYM2 model and the Poisson + RE model provide similar estimates; however the BYM2 model provides additional information
about the spatial vs. non-spatial structure of the variance.
