---
title: "Spatial Data Analysis in Python and R"
format:
  html:
    css: theming/quarto_styles.css 
    toc: true
    toc-location: left
---

## Example Dataset:  NYC traffic accidents, localized to US Census blocks

The dataset we're using is that used in the analysis published in 2019
[Bayesian Hierarchical Spatial Models: Implementing the Besag York Mollié Model in Stan](https://www.sciencedirect.com/science/article/pii/S1877584518301175).

The data consists of motor vehicle collisions in New York City,
as recorded by the NYC Department of Transportation, between the years 2005-2014,
restricted to collisions involving school age children 5-18 years of age as pedestrians.

Each crash was localized to the US Census tract in which it occurred, using boundaries from the 2010 United States Census,
using the [2010 Census block map for New York City](https://data.cityofnewyork.us/City-Government/2010-Census-Blocks/v2h8-6mxf)
The US Census tract is the lowest level of Census data resolution.
A tract contains between 2,500 and 8,000 residents and boundaries that follow visible features.
According to the [Census Bureau](https://www2.census.gov/geo/pdfs/reference/GARM/Ch10GARM.pdf)

> When first established, census tracts are to be as homogeneous as possible with respect to population characteristics, economic status, and living condition

### Measures

>We obtained motor vehicle collision data from the New York City
Department of Transportation for the ten most recent years of data
available at the time of request (2005-2014). Within this dataset, we
identified collisions involving school age children 5-18 years of age
as pedestrians.  We then assigned each crash to the census tract in
which it occurred, using boundaries from the 2010 United States
Census.

>We obtained 2010 US Census counts of youths aged 5-18 in each census
tract from the US Census Bureau [@us-census] We also obtained the
Census Bureau's American Community Survey (ACS) five-year estimates of
median household income and the percentage of commuters who traveled
to work by means other than a private vehicle (i.e., by walking,
bicycling, or using public transportation) for each tract for
2010-2014.  We constructed an index of social fragmentation based on
the work of Peter Congdon, using updated ACS estimates of vacant housing
units, householders living alone, non-owner occupied housing units,
and population having moved within the previous year.  We standardized
each of these metrics with a mean of zero and added them together as a
single index. Finally, we obtained street level annual average daily
traffic (AADT) data from the New York State Department of
Transportation on the New York Open Data portal.  We
created a spatial overlay of streets and census tracts to assign each
census tract the maximum AADT value of its underlying streets in 2015.

| Measures                                              | Median | Min    | Mean     | Max       |
|-------------------------------------------------------|--------|--------|----------|-----------|
| Youth pedestrian injuries, 2005-14                    | 6      | 0      | 8.2      | 57        |
| Population ages 5-18 years, 2010                      | 510    | 6      | 596.4    | 3,315     |
| Med. household income in USD, 2010-14                 | $53,890| $9,327 | $58,497  | $232,266  |
| Pct. commute by walk/cycle/public trans, 2010-14      | 73.9   | 9.7    | 69.8     | 100.0     |
| Standardized social fragmentation index               | -0.1   | -6.7   | 0.0      | 18.7      |
| Traffic Volume (AADT), 2015                           | 19,178 | 843    | 37,248   | 276,476   |


## Spatial Data manipulation in R and Python

Geographic information systems (GIS) data is any item which has a geographic location, either a single point or a set of bounding polygons.
Cartographic data (maps) are encoded as a set of records, one per map region.  The [shapefile format](https://en.wikipedia.org/wiki/Shapefile) is an open specification used to insure interoperability among GIS software packages.  When items in a dataset contain location labels, it is necessary to obtain a set of shapefiles for the corresponding map.
To compute with GIS data, we need packages which can do the geographic math.

- [GeoPandas](https://geopandas.org/en/stable/) extends the datatypes used by pandas to allow spatial operations on geometric types,
which are equivalent to the operations provided by R's [`sf`](https://r-spatial.github.io/sf/) package.
- [`libpysal`](https://pysal.org/libpysal/) is spatial analysis package which can analyze distance between locations; operations equivalent to R's [`spdep`](https://r-spatial.github.io/spdep/) package.

To visualize GIS data, we need to be able to draw maps and spatial graphs.

- In R, ggplot2 can produce beautiful maps for GIS data.
- The python equivalent of R's `ggplot2` is [`plotnine`](https://plotnine.org/).
- The [`splot`](https://pysal.org/notebooks/viz/splot/intro.html) provides additional plotting functions for spatial graph networks.


## Preliminary Data Analysis


### Notebook Startup: Load Packages / Libraries

```{r}
library(sf)
library(spdep) |> suppressPackageStartupMessages()
library(ggplot2)
library(tidyverse) |> suppressPackageStartupMessages()
```

### Load study data

The study data and GIS data have been assembled into a single GeoJSON file.

```{r}
nyc_geodata = st_read(file.path("data", "nyc_study.geojson"))
class(nyc_geodata)
names(nyc_geodata)
table(nyc_geodata$BoroName)
```


#### Traffic accidents - kids hit by cars

Plot the observed outcome `counts`.

*Note:* Count data is properly plotted as a bar chart, not histogram plus density, but the visualization is useful.

```{r}
ggplot(data = nyc_geodata, aes(x = count)) +
  geom_histogram(aes(y = after_stat(density)), bins = 58, color = 'DarkBlue', fill = "skyblue", alpha = 0.7) +
  geom_density(color = "red", linewidth = 1) +
  ggtitle("Distribution of accidents per tract") +
  xlab("accidents") +
  ylab("")
```



### Visualize by Map Areas

#### Show areal map, fill color is accidents per tract

```{r}
ggplot(data = nyc_geodata) +
geom_sf(aes(fill = count)) +
scale_fill_gradient(low = "lightblue", high = "red") +
labs(title = "NYC School-Age Traffic Accident Victims by Census Tract")
```

#### Show areal map, fill color is kid population per tract

```{r}
ggplot(data = nyc_geodata) +
    geom_sf(aes(fill = kid_pop)) +
    scale_fill_gradient(low = "lightblue", high = "red") +
    labs(title = "NYC Population Ages 5-18 Density by Census Tract")
```

#### Population: raw counts, density per square mile

```{r}
ggplot(data = nyc_geodata) +
    geom_sf(aes(fill = kids_per_sq_mi)) +
    scale_fill_gradient(low = "lightblue", high = "red") +
    labs(title = "NYC Population Ages 5-18 Density by Census Tract",
         subtitle = "Density measured as population per square mile")
```


##  Exercises

* Visualize `med_hh_inc` (median household income), `frag_index` (fragmentation index), and `traffic`.

* Create histograms of predictors `med_hh_inc`, `traffic`.   How can these be rescaled?

