---
title: "Implementing the ICAR Model in Stan"
format:
  html:
    css: theming/quarto_styles.css 
    syntax-definitions:
      - theming/stan.xml
    embed-resources: true
    toc: true
    toc-location: left
    grid:
      body-width: 1000px
---

## Notebook Setup

Import all libraries, load the NYC study data.

```{r}
library(sf)
library(spdep) |> suppressPackageStartupMessages()
library(ggplot2)
library(tidyverse) |> suppressPackageStartupMessages()
library(cmdstanr)
library(posterior)
library(bayesplot)
library(loo)
library(parallel)
library(igraph)
cores = floor(parallel::detectCores() / 2)


# helper function:  create edgelist for ICAR component
nbs_to_adjlist <- function(nb) {
    adj_matrix = nb2mat(nb,style="B")
    t(as_edgelist(graph_from_adjacency_matrix(adj_matrix, mode="undirected")))
}
```

```{r}
nyc_geodata = st_read("data/nyc_study.geojson")
names(nyc_geodata)
```

## Disconnected Components (and islands)

New York city consists of several islands; only the Bronx is part of the mainland; Brooklyn and Queens are part of Long Island, plus smaller islands City Island, Roosevelt Island, and the Rockaways.

```{r}
nyc_nbs = poly2nb(nyc_geodata, queen=FALSE)
summary(nyc_nbs)
```

*This is a problem for the ICAR model, which operates on a fully connected graph (single component)*

* For the NYC analysis paper, we hand edited the map of NYC (in R) to create a fully connected network graph.

* For this notebook, we restrict out attention to Brooklyn, the largest borough in NYC, which is a single network component.

```{r}
brooklyn_geodata = nyc_geodata[nyc_geodata$BoroName=='Brooklyn', ]
brooklyn_nbs = poly2nb(brooklyn_geodata, queen=FALSE)
brooklyn_coords = st_coordinates(st_centroid(brooklyn_geodata['geometry']))
plot(st_geometry(brooklyn_geodata), col='skyblue')
plot(brooklyn_nbs, coords=brooklyn_coords, add=TRUE, pch=20, cex=0.6)
```

```{r}
summary(brooklyn_nbs)
```

## Adding an ICAR Component to the Base Model

The ICAR model uses the spatial structure of the data, abstracted into a neighborhood network.

### Alternative encodings of neighborhood network

- $N \times N$ Adjacency matrix
  + Entries $(i,\ j)$ and $(j,\ i)$ are 1 when regions $n_i$ and $n_j$ are neighbors, 0 otherwise

- Undirected graph: regions are vertices, pairs of neighbors are edges
  + Encoded as *edges* - 2 column matrix, each row is a pair of neighbors $({n_i}, {n_j})$

The Stan ICAR model uses the latter.

### Stan ICAR Model Implementation

$$ p(\phi) \propto \exp \left\{ {- \frac{1}{2}} \sum_{i \sim j}{({\phi}_i - {\phi}_j)}^2 \right\} $$

Use Stan's vectorized operations to compute log probability density:
\fontsize{9pt}{9.4}\selectfont
```
   target += -0.5 * dot_self(phi[node1] - phi[node2]);
```
\normalsize
\vspace{0.2in}

Encode neighbor information as a graph.

* Nodes are indexed from 1:N.
* Edges indices are stored in a 2 x N array
  + each column is an edge
  + row 1: index of first node in edge pair, $n_i$
  + row 2: index of second node in edge pair, $n_j$

\fontsize{9pt}{9.4}\selectfont
```
  int<lower = 0> N;  // number of areal regions
  // spatial structure
  int<lower = 0> N_edges;  // number of neighbor pairs
  array[2, N_edges] int<lower = 1, upper = N> neighbors;  // node[1, j] adjacent to node[2, j]
```
\normalsize

## ICAR Model:  `poisson_icar.stan`

This file is in directory `stan/poisson_icar.stan`.

```{r}
poisson_icar_model_file = file.path('stan', 'poisson_icar.stan')
cat(readLines(poisson_icar_model_file), sep="\n")
```


## Data Prep

#### Get edgeset

- Compute this automatically from `nyc_geodata` spatial geometry component
  + Python package `libpysal`
  + R package `spdep`

```{r}
brooklyn_nbs_adj = nbs_to_adjlist(brooklyn_nbs)
brooklyn_nbs_adj[1:2, 1:10]
```

#### Assemble the input data 

```{r}
design_mat <- as.data.frame(brooklyn_geodata) %>%
  select(pct_pubtransit, med_hh_inc, traffic, frag_index) %>%
  mutate(med_hh_inc = log(med_hh_inc),
         traffic = log(traffic))
summary(design_mat)
```

```{r}
pois_icar_data <- list(
  N = nrow(brooklyn_geodata),
  y = as.integer(brooklyn_geodata$count),
  E = as.integer(brooklyn_geodata$kid_pop),
  K = 4,
  xs = design_mat,
  N_edges = ncol(brooklyn_nbs_adj),
  neighbors = brooklyn_nbs_adj
)
```

## Fitting the ICAR Model on the Brooklyn data

#### Model is compiled (as needed) on instantiation 

```{r}
pois_icar_mod = cmdstan_model(stan_file=poisson_icar_model_file)
```

#### Run the NUTS-HMC sampler, summarize results

```{r}
pois_icar_fit = pois_icar_mod$sample(data=pois_icar_data, parallel_chains=cores)
pois_icar_fit 
```

## Model Comparison: ICAR vs. ordinary random effects

```{r}
pois_re_mod = cmdstan_model(stan_file=file.path(
  'stan', 'poisson_re.stan')) 
pois_re_fit = pois_re_mod$sample(data=pois_icar_data, parallel_chains=cores)
pois_re_fit 
```

Which model provides a better fit (on the Brooklyn subset of the data)?

### Visual comparison

#### ICAR model

```{r}
y_rep <- as_draws_matrix(pois_icar_fit$draws("y_rep"))
ppc_dens_overlay(brooklyn_geodata$count, y_rep) +
                 ggtitle("Posterior Predictive Check: Poisson ICAR model")
```

#### RE model

```{r}
y_rep <- as_draws_matrix(pois_re_fit$draws("y_rep"))
ppc_dens_overlay(brooklyn_geodata$count, y_rep) +
                 ggtitle("Posterior Predictive Check: Poisson RE model")
```

### Leave-one-out cross-validation (LOO)

The Poisson-ICAR model falls into the category of "non-factorized models",
for which simple leave-one-out cross-validation fails to hold,
due to the fact that observations depend on other observations from 
different spatial units in addition to the dependence on the model parameters
Doing cross-validation would require recoding the model, and then re-fitting is $N$ times, once per omitted observation.

See paper: [Efficient leave-one-out cross-validation for Bayesian non-factorized normal and Student-t models](https://arxiv.org/abs/1810.10559),
which provides a thorough discussion of this problem and ways to evaluate non-factorized models,
and the LOO package vignette:
[Leave-one-out cross-validation for non-factorized models](https://mc-stan.org/loo/articles/loo2-non-factorized.html).

In these notebooks we will rely on visualizations and summary statistics.






