---
title: "The Intrinsic Conditional Auto-Regressive (ICAR) Model in Stan"
format:
  html:
    css: theming/quarto_styles.css 
    syntax-definitions:
      - theming/stan.xml
    embed-resources: true
    toc: true
    toc-location: left
    grid:
      body-width: 1000px
execute:
  eval: false
---

### ICAR Model Specification

* Conditional specification: multivariate normal random vector $\mathbf{\phi}$
where each ${\phi}_i$ is conditional on the values of its neighbors

* Joint specification rewrites to _Pairwise Difference_,
$$ p(\phi) \propto \exp \left\{ {- \frac{1}{2}} \sum_{i \sim j}{({\phi}_i - {\phi}_j)}^2 \right\} $$

* Each ${({\phi}_i - {\phi}_j)}^2$ contributes a
penalty term based on the distance between the values of neighboring regions

* $\phi$ is non-identifiable, constant added to $\phi$ washes out of ${\phi}_i - {\phi}_j$
    + A sum-to-zero constraint identifies and centers $\phi$

### Stan implementation

Encode neighbor information as graph edgeset, i.e. pairs of indices for neighbors $i, j$:
```stan
data {
  int<lower = 0> N;  // number of areal regions
  int<lower = 0> N_edges;  // number of neighbor pairs
  array[2, N_edges] int<lower = 1, upper = N> neighbors; // column-wise adjacent
  ...
```

Use Stan's `sum_to_zero_vector` to identify `phi`
```stan
parameters {
  sum_to_zero_vector[N] phi; // spatial random effects
  ...
```

Use Stan's vectorized operations to compute log probability density
```stan
model {
  target += -0.5 * dot_self(phi[node1] - phi[node2]);
  ...
```

## Notebook Setup

Import all libraries, load the NYC study data.

**Python**
```{python}
# import all libraries used in this notebook
import os
import numpy as np
import pandas as pd
import geopandas as gpd
import libpysal
import matplotlib
from splot.libpysal import plot_spatial_weights 
import plotnine as p9
%matplotlib inline

from cmdstanpy import CmdStanModel, cmdstan_path, cmdstan_version

import warnings
warnings.filterwarnings('ignore')
from utils_bym2 import get_scaling_factor, nbs_to_adjlist
from utils_dataviz import *
```
```{python}
nyc_geodata = gpd.read_file(os.path.join('data', 'nyc_study.geojson'))
nyc_geodata.columns
```

**R**
```{r}
library(sf)
library(spdep) |> suppressPackageStartupMessages()
library(ggplot2)
library(tidyverse) |> suppressPackageStartupMessages()
library(cmdstanr)
library(posterior)
library(bayesplot)
library(loo)
library(parallel)
cores = floor(parallel::detectCores() / 2)
source("utils_bym2.R")
source("utils_dataviz.R")
options(width = 120)
```

```{r}
nyc_geodata = st_read("data/nyc_study.geojson")
names(nyc_geodata)
```

## New York City Neighbor Graph

New York city consists of several islands; only the Bronx is part of the mainland;
Brooklyn and Queens are part of Long Island, Manhattan and Staten Island are large islands
with hundreds of census tracts, and
City Island, Roosevelt Island, and the Rockaways are smaller, comprised of 1, 2, and approximate 20
census tracts, respectively.

*This is a problem for the ICAR model, which operates on a fully connected graph (single component)*

**Python**
```{python}
w_nyc_nbs = libpysal.weights.Queen.from_dataframe(nyc_geodata, geom_col='geometry')
print(f'number of components: {w_nyc_nbs.n_components}')
print(f'islands? {w_nyc_nbs.islands}')
print(f'max number of neighbors per node: {w_nyc_nbs.max_neighbors}')
print(f'mean number of neighbors per node: {w_nyc_nbs.mean_neighbors}')
```

**R**
```{r}
nyc_nbs = poly2nb(nyc_geodata)
summary(nyc_nbs)
table(n.comp.nb(nyc_nbs)$comp.id)
```

* For the NYC analysis paper, we hand edited the map of NYC (in R) to create a fully connected network graph.

* For this notebook we restrict our attention to Brooklyn, the largest borough in NYC, which is a single network component.

**Python**
```{python}
brooklyn_geodata = nyc_geodata[nyc_geodata['BoroName']=='Brooklyn'].reset_index(drop=True)
brooklyn_nbs = libpysal.weights.Queen.from_dataframe(brooklyn_geodata, geom_col='geometry')
plot_spatial_weights(brooklyn_nbs, brooklyn_geodata) 
```

```{python}
print(f'number of components: {brooklyn_nbs.n_components}')
print(f'islands? {brooklyn_nbs.islands}')
print(f'max number of neighbors per node: {brooklyn_nbs.max_neighbors}')
print(f'mean number of neighbors per node: {brooklyn_nbs.mean_neighbors}')
```

**R**
```{r}
brooklyn_geodata = nyc_geodata[nyc_geodata$BoroName=='Brooklyn', ]
brooklyn_nbs = poly2nb(brooklyn_geodata)
brooklyn_coords = st_coordinates(st_centroid(brooklyn_geodata['geometry']))
plot(st_geometry(brooklyn_geodata), col='skyblue')
plot(brooklyn_nbs, coords=brooklyn_coords, add=TRUE, pch=20, cex=0.6)
```

```{r}
summary(brooklyn_nbs)
```

## ICAR Model:  `poisson_icar.stan`

### Coding the neighbor graph

- $N \times N$ Adjacency matrix
  + Entries $(i,\ j)$ and $(j,\ i)$ are 1 when regions $n_i$ and $n_j$ are neighbors, 0 otherwise

- Undirected graph: regions are vertices, pairs of neighbors are edges
  + Encoded as *edges* - 2 column matrix, each row is a pair of neighbors $({n_i}, {n_j})$

The Stan ICAR model uses the latter.
Encode neighbor information as a graph.

* Nodes are indexed from 1:N.
* Edges indices are stored in a 2 x N array
  + each column is an edge
  + row 1: index of first node in edge pair, $n_i$
  + row 2: index of second node in edge pair, $n_j$

```stan
  int<lower = 0> N;  // number of areal regions
  // spatial structure
  int<lower = 0> N_edges;  // number of neighbor pairs
  array[2, N_edges] int<lower = 1, upper = N> neighbors;  // node[1, j] adjacent to node[2, j]
```

### Coding the ICAR component

The joint distribution of the ICAR model is expressed as the _pairwise difference_
between neighboring regions in the neighbor graph.

$$ p(\phi) \propto \exp \left\{ {- \frac{1}{2}} \sum_{i \sim j}{({\phi}_i - {\phi}_j)}^2 \right\} $$

The joint distribution is non-identifiable;
adding any constant to all of the elements of $\phi$ leaves it unchanged,
therefore we add the constraint $\sum_{i} {\phi}_i = 0$.

* Use Stan's constrained parameter type `sum_to_zero_vector` to identify the joint distribution.

```stan
  sum_to_zero_vector[N] phi; // spatial effects
```


* Use Stan's vectorized operations to compute log probability density:

```stan
   target += -0.5 * dot_self(phi[node1] - phi[node2]);  // ICAR prior
```

### Putting it all together

This file is in directory `stan/poisson_icar.stan`.

**Python**
```{python}
poisson_icar_model_file = os.path.join('stan', 'poisson_icar.stan')

with open(poisson_icar_model_file, 'r') as file:
    contents = file.read()
    print(contents)
```

**R**
```{r}
poisson_icar_model_file = file.path('stan', 'poisson_icar.stan')
cat(readLines(poisson_icar_model_file), sep="\n")
```

## Data Prep

### Get edgeset

- Compute this automatically from `nyc_geodata` spatial geometry component
  + Python package `libpysal`
  + R package `spdep`

**Python**
```{python}
brooklyn_nbs_adj = nbs_to_adjlist(brooklyn_nbs)
brooklyn_nbs_adj
```

**R**
```{r}
brooklyn_nbs_adj = nbs_to_adjlist(brooklyn_nbs)
brooklyn_nbs_adj[1:2, 1:10]
```

### Assemble the input data 

**Python**
```{python}
design_vars = np.array(['pct_pubtransit','med_hh_inc', 'traffic', 'frag_index'])

design_mat = brooklyn_geodata[design_vars].to_numpy()
design_mat[:, 1] = np.log(design_mat[:, 1])
design_mat[:, 2] = np.log(design_mat[:, 2])

pd.DataFrame(data=design_mat).describe()
```

```{python}
pois_icar_data = {"N":brooklyn_geodata.shape[0],
             "y":brooklyn_geodata['count'].astype('int'),
             "E":brooklyn_geodata['kid_pop'].astype('int'),
             "K":4,
             "xs":design_mat,
             "N_edges": brooklyn_nbs_adj.shape[1],
             "neighbors": brooklyn_nbs_adj
}
```

**R**
```{r}
design_mat <- as.data.frame(brooklyn_geodata) %>%
  select(pct_pubtransit, med_hh_inc, traffic, frag_index) %>%
  mutate(med_hh_inc = log(med_hh_inc),
         traffic = log(traffic))
summary(design_mat)
```

```{r}
pois_icar_data <- list(
  N = nrow(brooklyn_geodata),
  y = as.integer(brooklyn_geodata$count),
  E = as.integer(brooklyn_geodata$kid_pop),
  K = 4,
  xs = design_mat,
  N_edges = ncol(brooklyn_nbs_adj),
  neighbors = brooklyn_nbs_adj
)
```

## Fitting the ICAR Model on the Brooklyn data

### Model is compiled (as needed) on instantiation 

**Python**
```{python}
pois_icar_mod = CmdStanModel(stan_file=poisson_icar_model_file)
```

**R**
```{r}
pois_icar_mod = cmdstan_model(stan_file=poisson_icar_model_file)
```

### Run the NUTS-HMC sampler, summarize results

**Python**
```{python}
pois_icar_fit = pois_icar_mod.sample(data=pois_icar_data)
pois_icar_fit.summary().round(2).loc[
  ['beta_intercept', 'beta0', 'betas[1]', 'betas[2]', 'betas[3]', 'betas[4]', 'sigma']]
```

**R**
```{r}
pois_icar_fit = pois_icar_mod$sample(data=pois_icar_data, parallel_chains=cores, refresh=0)
pois_icar_fit 
```


## Model Checking

### Visualizing the ICAR component

To see the spatial structure of the data, we plot the covariance matrix of `phi`.
We expect neighboring regions to be correlated, and distant regions to be uncorrelated.
To do this, we have written a helper function, `plot_icar_corr_matrix`,
which plots just the upper half of the correlation matrix of `phi`,
with red indicating positive correlation and blue indicating negative correlation.
The census tracts are listed roughly in geographic order, from north-to-south and
east-to-west, thus adjacent elements along the diagonal are spatially adjacent.

**Python**
```{python}
phi_icar = pois_icar_fit.stan_variable("phi")
corr_plot = plot_icar_corr_matrix(phi_icar, 'ICAR correlation matrix', (10, 10))
corr_plot
```

**R**
```{r}
phi_icar <- as_draws_matrix(pois_icar_fit$draws("phi"))
plot_icar_corr_matrix(phi_icar, 'ICAR correlation matrix')
```

The correlation matrix shows strong correlation between neighbors, along the diagonal,
and low correlation between distant regions, in the upper left-hand corner.

### Posterior predictive checks

**Python**
```{python}
y_rep_icar = pois_icar_fit.stan_variable("y_rep")
print(ppc_central_interval(y_rep_icar, pois_icar_data['y']))

ppc_plot = ppc_y_yrep_overlay(y_rep_icar, pois_icar_data['y'],
                                'Poisson + ICAR model PPC\ny (blue dot) vs. y_rep (orange 50% central interval, grey full extent)')
ppc_plot
```

**R**
```{r}
y_rep_icar <- as_draws_matrix(pois_icar_fit$draws("y_rep"))
ppc_central_interval(y_rep_icar, pois_icar_data$y)
ppc_y_yrep_overlay(y_rep_icar, pois_icar_data$y,
                     'Poisson + ICAR model PPC\ny (blue dot) vs. y_rep (orange 50% central interval, grey full extent)')
```

## Model Comparison: Poisson + ICAR vs. Poisson vs. Poisson + RE

The Poisson-ICAR model falls into the category of "non-factorized models",
for which simple leave-one-out cross-validation fails to hold,
due to the fact that observations depend on other observations from 
different spatial units in addition to the dependence on the model parameters
Doing cross-validation would require recoding the model, and then re-fitting is $N$ times, once per omitted observation.
Therefore, we compare summary statistics, visualizations, and model runtimes.

#### Poisson model (with centered predictors)

**Python**
```{python}
pois_mod = CmdStanModel(stan_file=os.path.join(
  'stan', 'poisson_ctr_preds.stan'))
pois_fit = pois_mod.sample(data=pois_icar_data)
pois_fit.summary().round(2).loc[
  ['beta_intercept', 'beta0', 'betas[1]', 'betas[2]', 'betas[3]', 'betas[4]']]
```

**R**
```{r}
pois_re_mod = cmdstan_model(stan_file=file.path(
  'stan', 'poisson_ctr_preds.stan')) 
pois_re_fit = pois_re_mod$sample(data=pois_icar_data, parallel_chains=cores, refresh=0)
pois_re_fit 
```

#### Poisson + RE model

**Python**
```{python}
pois_re_mod = CmdStanModel(stan_file=os.path.join(
  'stan', 'poisson_re.stan'))
pois_re_fit = pois_re_mod.sample(data=pois_icar_data)
pois_re_fit.summary().round(2).loc[
  ['beta_intercept', 'beta0', 'betas[1]', 'betas[2]', 'betas[3]', 'betas[4]', 'sigma']]
```

**R**
```{r}
pois_re_mod = cmdstan_model(stan_file=file.path(
  'stan', 'poisson_re.stan')) 
pois_re_fit = pois_re_mod$sample(data=pois_icar_data, parallel_chains=cores, refresh=0)
pois_re_fit 
```

### Compare Posterior predictive check coverage

**Python**
```{python}
print(ppc_central_interval(y_rep_icar, pois_icar_data['y']))

y_rep_pois = pois_fit.stan_variable("y_rep")
print(ppc_central_interval(y_rep_pois, pois_icar_data['y']))

y_rep_re = pois_re_fit.stan_variable("y_rep")
print(ppc_central_interval(y_rep_re, pois_icar_data['y']))
```

**R**
```{r}
y_rep_icar <- as_draws_matrix(pois_icar_fit$draws("y_rep"))
ppc_central_interval(y_rep_icar, pois_icar_data$y)

y_rep_pois <- as_draws_matrix(pois_re_fit$draws("y_rep"))
ppc_central_interval(y_rep_pois, pois_icar_data$y)

y_rep_re <- as_draws_matrix(pois_re_fit$draws("y_rep"))
ppc_central_interval(y_rep_re, pois_icar_data$y)
```


### Poisson Model PPC plots

**Python**
```{python}
ppc_plot = ppc_y_yrep_overlay(y_rep_pois, pois_icar_data['y'],
                                'Poisson model PPC\ny (blue dot) vs. y_rep (orange 50% central interval, grey full extent)')
ppc_plot
```

**R**
```{r}
ppc_y_yrep_overlay(y_rep_pois, pois_icar_data$y,
                     'Poisson model PPC\ny (blue dot) vs. y_rep (orange 50% central interval, grey full extent)')
```

### Poisson + RE Model PPC plots

**Python**
```{python}
ppc_plot = ppc_y_yrep_overlay(y_rep_re, pois_icar_data['y'],
                                'Poisson + RE model PPC\ny (blue dot) vs. y_rep (orange 50% central interval, grey full extent)')
ppc_plot
```

**R**
```{r}
ppc_y_yrep_overlay(y_rep_re, pois_icar_data$y,
                     'Poisson + RE model PPC\ny (blue dot) vs. y_rep (orange 50% central interval, grey full extent)')
```

### Poisson + ICAR Model PPC plots

**Python**
```{python}
ppc_plot = ppc_y_yrep_overlay(y_rep_icar, pois_icar_data['y'],
                                'Poisson + ICAR model PPC\ny (blue dot) vs. y_rep (orange 50% central interval, grey full extent)')
ppc_plot
```

**R**
```{r}
ppc_y_yrep_overlay(y_rep_icar, pois_icar_data$y,
                     'Poisson + ICAR model PPC\ny (blue dot) vs. y_rep (orange 50% central interval, grey full extent)')
```


## Discussion

The Poisson + RE model has somewhat better predictive coverage because the random effects component `theta` effectively accounts for the overdispersed nature of the data.   However, visualization of the ICAR component demonstrates that the ICAR component does a good job of capturing the spatial structure of the data
and also provides good predictive performance.
In the next notebook, we show how to combine both components into the BYM2 model.
