---
title: "Implementing the BYM2 Model in Stan"
format:
  html:
    css: theming/quarto_styles.css 
    syntax-definitions:
      - theming/stan.xml
    embed-resources: true
    toc: true
    toc-location: left
    grid:
      body-width: 1000px
---

## Notebook Setup

Import all libraries, load the NYC study data.

```{r}
library(sf)
library(spdep) |> suppressPackageStartupMessages()
library(ggplot2)
library(tidyverse) |> suppressPackageStartupMessages()
library(cmdstanr)
library(posterior)
library(bayesplot)
library(loo)
library(parallel)
library(igraph)
cores = floor(parallel::detectCores() / 2)


# helper function:  create edgelist for ICAR component
nbs_to_adjlist <- function(nb) {
    adj_matrix = nb2mat(nb,style="B")
    t(as_edgelist(graph_from_adjacency_matrix(adj_matrix, mode="undirected")))
}
```

```{r}
nyc_geodata = st_read("data/nyc_study.geojson")
names(nyc_geodata)
```

## Disconnected Components (and islands)

New York city consists of several islands; only the Bronx is part of the mainland; Brooklyn and Queens are part of Long Island, plus smaller islands City Island, Roosevelt Island, and the Rockaways.

*This is a problem for the ICAR model, which operates on a fully connected graph (single component)*

* For the NYC analysis paper, we hand edited the map of NYC (in R) to create a fully connected network graph.

* For this notebook, we will restrict out attention to Brooklyn, the largest borough in NYC, which is a single network component.


```{r}
brooklyn_geodata = nyc_geodata[nyc_geodata$BoroName=='Brooklyn', ]
brooklyn_nbs = poly2nb(brooklyn_geodata, queen=FALSE)
brooklyn_coords = st_coordinates(st_centroid(brooklyn_geodata['geometry']))
plot(st_geometry(brooklyn_geodata), col='skyblue')
plot(brooklyn_nbs, coords=brooklyn_coords, add=TRUE, pch=20, cex=0.6)
```

```{r}
summary(brooklyn_nbs)
```


## From ICAR to BYM2

* Combines both ICAR component $\phi$ and ordinary random effects $\theta$ as
$$\left( (\sqrt{\, {\rho} / s}\, \ )\,\phi^* + (\sqrt{1-\rho})\,\theta^* \right) \sigma $$

* Parameter $\rho$ answers the question:  how much of the observed variance is spatial?

* Don't need to run analysis, e.g. Moran's I - the model sorts it out for you.

## BYM2 Model:  `bym2.stan`

This file is in directory `stan/bym2.stan`.

```{r}
bym2_model_file = file.path('stan', 'bym2.stan')
cat(readLines(bym2_model_file), sep="\n")
```

## Data Prep

### Get edgeset

- Compute this automatically from `nyc_geodata` spatial geometry component
  + Python package `libpysal`
  + R package `spdep`

```{r}
brooklyn_nbs_adj = nbs_to_adjlist(brooklyn_nbs)
brooklyn_nbs_adj[1:2, 1:10]
```

### Compute scaling factor `tau`

Computed in R:  value 0.658

R Script:
```r
# computes the inverse of a sparse precision matrix
# sub-optimal implementation - better to use INLA
q_inv_dense <- function(Q, A = NULL) {
  Sigma <- Matrix::solve(Q)   ## need sparse matrix solver
  if (is.null(A))
    return(Sigma)
  else {
    A <- matrix(1,1, nrow(Sigma))
    W <- Sigma %*% t(A)
    Sigma_const <- Sigma - W %*% solve(A %*% W) %*% t(W)
    return(Sigma_const)
  }
}

get_scaling_factor = function(adj_list) {
    N = ncol(adj_list)
    # Build the adjacency matrix using edgelist
    adj_matrix = sparseMatrix(i=adj_list[1, ], j=adj_list[2, ], x=1, symmetric=TRUE)

    # Create ICAR precision matrix  (diag - adjacency): this is singular
    Q =  Diagonal(N, rowSums(adj_matrix)) - adj_matrix
    # Add a small jitter to the diagonal for numerical stability (optional but recommended)
    Q_pert = Q + Diagonal(N * max(diag(Q)) * sqrt(.Machine$double.eps)

    # Compute the diagonal elements of the covariance matrix
    Q_inv = q_inv_dense(Q_pert, adj_matrix)

    # Compute the geometric mean of the variances, which are on the diagonal of Q.inv
    return(exp(mean(log(diag(Q_inv)))))
}
```

#### Assemble the input data 

```{r}
design_mat <- as.data.frame(brooklyn_geodata) %>%
  select(pct_pubtransit, med_hh_inc, traffic, frag_index) %>%
  mutate(med_hh_inc = log(med_hh_inc),
         traffic = log(traffic))
summary(design_mat)
```

```{r}
bym2_data <- list(
  N = nrow(brooklyn_geodata),
  y = as.integer(brooklyn_geodata$count),
  E = as.integer(brooklyn_geodata$kid_pop),
  K = 4,
  xs = design_mat,
  N_edges = ncol(brooklyn_nbs_adj),
  neighbors = brooklyn_nbs_adj,
  tau = 0.658
)
```

## Fitting the BYM2 Model on the Brooklyn data

#### Model is compiled (as needed) on instantiation 

```{r}
bym2_mod = cmdstan_model(stan_file=bym2_model_file)
```

#### Run the NUTS-HMC sampler, summarize results

```{r}
bym2_fit = bym2_mod$sample(data=bym2_data, parallel_chains=cores)
bym2_fit 
```

## Model Comparison: BYM2 vs ICAR vs. ordinary random effects


#### ICAR model

```{r}
pois_icar_mod = cmdstan_model(stan_file=file.path(
  'stan', 'poisson_icar.stan')) 
pois_icar_fit = pois_icar_mod$sample(data=bym2_data, parallel_chains=cores)
pois_icar_fit 
```

#### Ordinary random effects model

```{r}
pois_re_mod = cmdstan_model(stan_file=file.path(
  'stan', 'poisson_re.stan')) 
pois_re_fit = pois_re_mod$sample(data=bym2_data, parallel_chains=cores)
pois_re_fit 
```

Which model provides a better fit (on the Brooklyn subset of the data)?

### Visual comparison

#### BYM2 model

```{r}
y_rep <- as_draws_matrix(bym2_fit$draws("y_rep"))
ppc_dens_overlay(brooklyn_geodata$count, y_rep) +
                 ggtitle("Posterior Predictive Check: BYM2 model")
```


#### ICAR model

```{r}
y_rep <- as_draws_matrix(pois_icar_fit$draws("y_rep"))
ppc_dens_overlay(brooklyn_geodata$count, y_rep) +
                 ggtitle("Posterior Predictive Check: Poisson ICAR model")
```

#### RE model

```{r}
y_rep <- as_draws_matrix(pois_re_fit$draws("y_rep"))
ppc_dens_overlay(brooklyn_geodata$count, y_rep) +
                 ggtitle("Posterior Predictive Check: Poisson RE model")
```

### Leave-one-out cross-validation (LOO)

```{r}
loo_bym2 <- loo(bym2_fit$draws("log_lik"), save_psis = TRUE)
loo_icar_pois <- loo(pois_icar_fit$draws("log_lik"), save_psis = TRUE)
loo_re_pois <- loo(pois_re_fit$draws("log_lik"), save_psis = TRUE)

loo_compare(loo_bym2, loo_icar_pois, loo_re_pois)
```
