---
title: "Extending the BYM2 Model for Disconnected Graphs"
format:
  html:
    css: theming/quarto_styles.css 
    syntax-definitions:
      - theming/stan.xml
    embed-resources: true
    toc: true
    toc-location: left
    grid:
      body-width: 1000px
execute:
  eval: false
  keep-ipynb: true
---


## Notebook setup

### Libraries and helper functions.

**Python**
```{python}
# import all libraries used in this notebook
import os
import numpy as np
import pandas as pd
import geopandas as gpd
import libpysal
import matplotlib
from splot.libpysal import plot_spatial_weights 
%matplotlib inline

import warnings
warnings.filterwarnings('ignore')

from cmdstanpy import CmdStanModel, cmdstan_path, cmdstan_version

from utils_bym2 import get_scaling_factor, nbs_to_adjlist
from utils_nyc_map import *
from utils_dataviz import *
```

**R**
```{r}
library(ggplot2)
library(igraph)
library(Matrix)
library(parallel)
library(sf)
library(spdep) |> suppressPackageStartupMessages()
library(tidyverse) |> suppressPackageStartupMessages()

library(cmdstanr)
library(posterior)
library(bayesplot)
library(loo)
cores = floor(parallel::detectCores() / 2)
source("utils_bym2.R")
source("utils_dataviz.R")
source("utils_nyc_map.R")
```

## The BYM2 Model for Disconnected Graphs: Freni-Sterrantino et al, 2018

## A quick review of neighbor graphs

In the context of an areal map, the neighbor graph is used to model spatial relationships between
regions (such as states, districts, or census tracts). Each region is represented as a node, and an
edge is drawn between two nodes if their corresponding regions share a common boundary.
Edges have no direction, i.e., an edge between $i$ and $j$ implies
that $i$ is a neighbor of $j$ and $j$ is a neighbor of $i$.

*Components* represent groups of regions that are connected through shared boundaries, either
directly or indirectly.  The size of a component is the number of nodes in it.  An island, or
singleton, is a component of size 1; in this notebook we distinguish between *connected components*
of size 2 or greater and *singletons*.

If it is possible to find a path (series of edges) which connect every region in the graph to every
other region in the graph, then the graph is *fully connected*, i.e., it consists of a single
component.  For example, at the state level, the neighbor graph of the 48 continental states of the
U.S. is fully connected, but the graph of all 50 U.S. states and territories is *disconnected*;
it consists of a connected component and several islands.

## The BYM2 Model for Disconnected Graphs: Freni-Sterrantino et al, 2018

The ICAR model requires that the neighbor graph is fully connected for two reasons:

* The joint distribution is computed from the pairwise differences between a node and its neighbors;
singleton nodes have no neighbors and are therefore undefined.

* Even if the graph doesn't have any singleton nodes, when the graph has multiple connected components
a sum-to-zero constraint on the entire vector fails to properly identify the model.

As the census tract map of NYC shows,
if the metric used to compute spatial adjacency is direct overland adjacency,
then the neighbor graph of New York City is has several connected components and a few singleton nodes.

In order to apply the BYM2 model to the full NYC dataset, it is necessary to either:

* Edit the neighbor graph to create a fully connected single component for NYC, or
* Extend the BYM2 model to account for disconnected components and singleton nodes.

For the analysis in Morris et al, 2019, we edited the neighbor graph by adding
enough edges to create a single connected component.

### Freni-Sterrantion recommendations

In this notebook, we instead extend the BYM2 model to account for disconnected graphs and islands,
following the recommendations from
[A note on intrinsic Conditional Autoregressive models for disconnected graphs](https://arxiv.org/abs/1705.04854),
Freni-Sterrantino et.al. 2018. 

* Component nodes are given the BYM2 prior
* Singleton nodes (islands) are given a standard Normal prior
* Compute per-connected component scaling factor
* Impose a sum-to-zero constraint on each connected component


## Stan Implementation: `bym2_multicomp.stan`

The BYM2_multicomp model requires more additional data inputs: the number of components, their size,
scaling factor, and the number of singleton nodes.  Because neighbor graph uses the spatial
dataframe row indices as the node labels, it is critical that the order of all data inputs matches
the dataframe ordering, i.e., that for any row $i$ in the spatial dataframe, the $i^{th}$ element of
the observed outcome `y` and design matrix `xs` match that row, and the node ids used in the
`neighbors` edge pairs array.  Data preprocessing consists of the following operations, in order:

* Create the neighbor graph from the spatial dataframe; add graph component labels to the dataframe.

* Create mapping from component labels to component sizes.

* Reorder the spatial dataframe so that it can be sliced by component, with singleton nodes last,
by using the component id as the sort key and the component size, descending, as the sort order.

* Rebuild the neighbor graph so the its indices line up with the spatial dataframe row index.

We have written a series of helper functions in R and Python to do this
which are in files `utils_nyc_map.py` and `utils_nyc_map.R`.
For further details on editing spatial data structures see the notebook on spatial data:
[Spatial Data Analysis in Python and R](https://github.com/mitzimorris/geomed_2024/blob/main/h2_spatial_data.qmd).

By reordering the spatial data frame by component membership, with singleton nodes at the end of the
dataframe, we can easily implement both the component-wise sum to zero constraint and compute the
spatial effects for connected nodes, using Stan's
[range indexing](https://mc-stan.org/docs/reference-manual/expressions.html#language-multi-indexing.section)
expressions, which return slices of an array.  See the Stan User's Guide, section
[slicing with range indexes](https://mc-stan.org/docs/stan-users-guide/multi-indexing.html#slicing-with-range-indexes)
for details.

### Changes to the `data` block

The model needs a few more pieces of information:
the number of connected components, the size and scaling factor for each.

```stan
data {
  ...
  int<lower=0, upper=N> N_components;
  array[N_components] int<lower=1, upper=N> component_sizes;
  vector<lower=0>[N_components] scaling_factors;  // vector instead of a scalar

  // neighbor graph structure
  int<lower = 0> N_edges;  // number of neighbor pairs
  array[2, N_edges] int<lower = 1, upper = (sum(component_sizes))> neighbors;  // was upper = N
}
```

### Changes to the `transformed data` block

First we check that the input data is consistent.

```stan
transformed data {
  ...
  int N_connected = sum(component_sizes);
  int N_singletons = N - N_connected;
  if (N_singletons < 0) {
    reject("Inconsistent inputs: sum(component_sizes) > N");
  }
  ...
```

Then we use component sizes to compute the begin and end indices needed to slice the spatial vector `phi`.
At the same time, we create vector `taus`, the per-region scaling factor.

```stan
  for (n in 1:N_components) {
    taus[idx: idx + component_sizes[n] - 1]
      = rep_array(scaling_factors[n], idx, component_sizes[n]);
    component_idxs[n, 1] = idx;
    component_idxs[n, 2] = idx + component_sizes[n] - 1;
    idx += component_sizes[n];
  }
```

### Changes to the `parameters` block:  constraining transforms

In the BYM2 model for a fully connected graph the sum-to-zero constraint on `phi`
is implemented directly by declaring `phi` to be a `sum_to_zero_vector`, which is a
[constrained parameter type](https://mc-stan.org/docs/reference-manual/transforms.html#variable-transforms.chapter).
The declaration:

```stan
  sum_to_zero_vector[N] phi;  // spatial effects
```

creates a *constrained* variable of length $N$, with a corresponding unconstrained variable of length $N-1$.

For the BYM2_multicomp model, we need to do declare the *unconstrained* parameter vector `phi_raw`
and the constraining transform, which is applied component-wise, to slices of `phi_raw`.
The number of connected components is $N\_components$, therefore the length of vector `phi_raw` is
$N - N\_components$.

```stan
  vector[N - N_components] phi_raw;  // spatial effects
```

In the `transformed parameters` block, we apply the constraining transform.

```stan
  vector[N_connected] phi = zero_sum_components_lp(phi_raw, component_idxs, component_sizes);
```

In the `functions` block, we define two functions

* `zero_sum_constrain_lp`: the constraining transform, following the `zero_sum_vector` implementation.

```stan
  /**
   * Constrain sum-to-zero vector
   *
   * @param y unconstrained zero-sum parameters
   * @return vector z, the vector whose slices sum to zero
   */
  vector zero_sum_constrain_lp(vector y) {
    int N = num_elements(y);
    vector[N + 1] z = zeros_vector(N + 1);
    real sum_w = 0;
    for (ii in 1:N) {
      int i = N - ii + 1; 
      real n = i;
      real w = y[i] * inv_sqrt(n * (n + 1));
      sum_w += w;
      z[i] += sum_w;     
      z[i + 1] -= w * n;    
    }
    return z;
  }
```

* `zero_sum_components_lp`: slices vector `phi` by component, applies constraining transform to each.

```stan
functions {
  /**
   * Component-wise constrain sum-to-zero vectors
   *
   * @param phi unconstrained vector of zero-sum slices
   * @param idxs component start and end indices
   * @param sizes component sizes
   * @return vector phi_ozs, the vector whose slices sum to zero
   */
  vector zero_sum_components_lp(vector phi,
                                array[ , ] int idxs,
                                array[] int sizes) {
    vector[sum(sizes)] phi_ozs;
    int idx_phi = 1;
    int idx_ozs = 1;
    for (i in 1:size(sizes)) {
      phi_ozs[idx_ozs : idx_ozs + sizes[i] - 1] =
        zero_sum_constrain_lp(segment(phi, idx_phi, sizes[i] - 1));
      idx_phi += sizes[i] - 1;
      idx_ozs += sizes[i];
    }
    return phi_ozs;
  }
```

### Putting it all together:  `bym2_multicomp.stan`

**Python**
```{python}
bym2_multicomp_file = os.path.join('stan', 'bym2_multicomp.stan')

with open(bym2_multicomp_file, 'r') as file:
    contents = file.read()
    print(contents)
```

**R**
```{r}
bym2_multicomp_file = file.path('stan', 'bym2_multicomp.stan')
cat(readLines(bym2_multicomp_file), sep="\n")
```


## Fitting the BYM2_multicomp Model to the NYC Dataset

### Data Assembly

We need to assemble a dictionary or list for all variables declared in the model's data block.
These are:

```stan
data {
  int<lower=0> N;
  array[N] int<lower=0> y; // count outcomes
  vector<lower=0>[N] E; // exposure
  int<lower=1> K; // num covariates
  matrix[N, K] xs; // design matrix

  int<lower=0, upper=N> N_components;
  array[N_components] int<lower=1, upper=N> component_sizes;
  vector<lower=0>[N_components] scaling_factors;
  int<lower=0, upper=N> N_singletons;

  // neighbor graph structure
  int<lower = 0> N_edges;  // number of neighbor pairs
  array[2, N_edges] int<lower = 1, upper = (N - N_singletons)> neighbors;  // columnwise adjacent
}
```

#### Load the NYC dataset

**Python**
```{python}
nyc_geodata = gpd.read_file(os.path.join('data', 'nyc_study.geojson'))
nyc_geodata.columns
```

**R**
```{r}
nyc_geodata = st_read("data/nyc_study.geojson")
names(nyc_geodata)
```

#### NYC neighbor graph cleanup

Create the neighbors graph and plot the map.  This shows a set of spurious connections
between Manhattan and Brooklyn and Queens, (likely due to bridges and tunnels).


**Python**
```{python}
nyc_nbs_all = libpysal.weights.Queen.from_dataframe(nyc_geodata, geom_col='geometry')
plot_spatial_weights(nyc_nbs_all, nyc_geodata)
```

**R**
```{r}
nyc_nbs_all = poly2nb(nyc_geodata)
nyc_coords = st_coordinates(st_centroid(nyc_geodata['geometry']))
plot(st_geometry(nyc_geodata), col='skyblue')
plot(nyc_nbs_all, coords=nyc_coords, add=TRUE, pch=20, cex=0.6)
```

In order to ensure that the spatial graph accurately reflects real‐world connectivity,
we have written a helper function to disconnect Manhattan
from Queens and Brooklyn.  However, as Queens and Brooklyn are in fact the same land mass
(Long Island), we don't try to disconnect Brooklyn from Queens.

To check that the neighbor graph is correct, we replot it.

**Python**
```{python}
nyc_nbs = nyc_cleanup(nyc_nbs_all, nyc_geodata)
plot_spatial_weights(nyc_nbs, nyc_geodata)
```

**R**
```{r}
nyc_components = nyc_sort_by_comp_size(nyc_geodata)
nyc_nbs = nyc_components$nb
nyc_gdf = nyc_components$sf_sorted
sizes = nyc_components$component_sizes

nyc_coords = st_coordinates(st_centroid(nyc_gdf$geometry[1:length(nyc_nbs)]))
plot(st_geometry(nyc_gdf), col='skyblue')
plot(nyc_nbs, coords=nyc_coords, add=TRUE, pch=20, cex=0.6)
```

#### Arranging the dataset by component

To reorder the spatial dataframe so that it can be sliced by component, with singleton nodes last,
and rebuild the corresponding neighbor graph, we have written function `nyc_sort_by_comp_size`,
which takes into account the necessity of correcting for problems in the NYC geodata.
A general version for maps where the spatial graph lines up with real-world connectivity,
this function could be simplified.

**Python**
```{python}
(nyc_nbs, nyc_gdf, sizes) = nyc_sort_by_comp_size(nyc_geodata)
plot_spatial_weights(nyc_nbs, nyc_gdf)
```
```{python}
nyc_gdf[['NTAName', 'comp_id', 'comp_size']].head(4)
```
```{python}
nyc_gdf[['NTAName', 'comp_id', 'comp_size']].tail(4)
```

**R**
```{r}
nyc_components = nyc_sort_by_comp_size(nyc_geodata)
nyc_nbs = nyc_components$nb
nyc_gdf = nyc_components$sf_sorted
sizes = nyc_components$component_sizes

nyc_coords = st_coordinates(st_centroid(nyc_gdf['geometry']))
plot(st_geometry(nyc_gdf), col='skyblue')
plot(nyc_nbs, coords=nyc_coords, add=TRUE, pch=20, cex=0.6)
```

#### Get edgeset

**Python**
```{python}
nyc_nbs_adj = nbs_to_adjlist(nyc_nbs)
nyc_nbs_adj
```

**R**
```{r}
nyc_nbs_adj = nbs_to_adjlist(nyc_nbs)
nyc_nbs_adj[1:2, 1:10]
```

#### Get regression inputs

**Python**
```{python}
N = nyc_gdf.shape[0]
y = nyc_gdf['count'].astype('int')
E = nyc_gdf['kid_pop'].astype('int')
K = 4

design_vars = np.array(['pct_pubtransit','med_hh_inc', 'traffic', 'frag_index'])
design_mat = nyc_gdf[design_vars].to_numpy()
design_mat[:, 1] = np.log(design_mat[:, 1])
design_mat[:, 2] = np.log(design_mat[:, 2])
pd.DataFrame(data=design_mat).describe()
```

**R**
```{r}
N = nrow(nyc_gdf)
y = as.integer(nyc_gdf$count)
E = as.integer(nyc_gdf$kid_pop)
K = 4

design_mat <- as.data.frame(nyc_gdf) %>%
  select(pct_pubtransit, med_hh_inc, traffic, frag_index) %>%
  mutate(med_hh_inc = log(med_hh_inc),
         traffic = log(traffic))
summary(design_mat)
```

#### Compute number of components, sizes, and number of singleton nodes

**Python**
```{python}
component_sizes = [x for x in sizes if x > 1]
N_components = len(component_sizes)
print("N_components ", N_components, " component_sizes ", component_sizes)
print("N connected ", np.sum(component_sizes))
```

**R**
```{r}
component_sizes <- sizes[sizes > 1]  # Filter sizes greater than 1
N_components <- length(component_sizes)
cat("N_components:", N_components, "component_sizes:", component_sizes, "\n")
cat("N connected:", sum(component_sizes), "\n")
```

#### Compute per-component scaling factors

**Python**
```{python}
scaling_factors = np.ones(N_components)
for i in range(N_components):
    comp_gdf = nyc_gdf[nyc_gdf['comp_id'] == i].reset_index(drop=True)
    comp_nbs = libpysal.weights.Queen.from_dataframe(comp_gdf, geom_col='geometry')
    component_w = libpysal.weights.W(comp_nbs.neighbors, comp_nbs.weights)
    scaling_factors[i] = get_scaling_factor(component_w)

print(scaling_factors)
```

**R**
```{r}
scaling_factors <- rep(1, N_components)  # Initialize vector of ones

for (i in seq_len(N_components)) {
    comp_gdf <- nyc_gdf[nyc_gdf$comp_id == i, ]  # Select component i
    comp_nbs <- poly2nb(comp_gdf)  # Create Queen contiguity neighbors
    scaling_factors[i] <- get_scaling_factor(comp_nbs)  # Compute scaling factor
}

print(scaling_factors)
```

#### Assemble the input data into a Python dict, R list

**Python**
```{python}
# assemble nyc_data_dict
bym2_data = {
    'N':N,
    'y':y,
    'E':E,
    'K':K,
    'xs':design_mat,
    'N_components':N_components,
    'component_sizes': component_sizes,
    'N_edges':nyc_nbs_adj.shape[1],
    'neighbors':nyc_nbs_adj,
    'scaling_factors': scaling_factors
}
```

**R**
```{r}
bym2_data <- list(
  N = N,
  y = y,
  E = E,
  K = 4,
  xs = design_mat,
  N_components = N_components,
  component_sizes = component_sizes,
  N_edges = ncol(nyc_nbs_adj),
  neighbors = nyc_nbs_adj,
  scaling_factors = scaling_factors
)
```

### Model compilation

**Python**
```{python}
bym2_multicomp_mod = CmdStanModel(stan_file=bym2_multicomp_file)
```

**R**
```{r}
bym2_multicomp_mod = CmdStanModel(bym2_multicomp_file)
```

### Run the NUTS-HMC sampler, summarize results

**Python**
```{python}
bym2_multicomp_fit = bym2_multicomp_mod.sample(data=bym2_data, iter_warmup=3000, iter_sampling=2000)

bym2_multicomp_summary = bym2_multicomp_fit.summary()
bym2_multicomp_summary.round(2).loc[
  ['beta_intercept', 'beta0', 'betas[1]', 'betas[2]', 'betas[3]', 'betas[4]', 'sigma', 'rho']]
```

**R**
```{r}
bym2_multicomp_fit = bym2_multicomp_mod$sample(data=bym2_data, parallel_chains=cores,
		   iter_warmup=3000, iter_sampling=2000, refresh=0)
bym2_multicomp_fit$print(variables = c('beta_intercept', 'beta0', 'betas[1]', 'betas[2]', 'betas[3]', 'betas[4]', 'sigma', 'rho')) 
```

## Model Checking

### Visualizing the ICAR component

To see the spatial structure of the data, we plot the covariance matrix of `phi`.
We expect neighboring regions to be correlated, and distant regions to be uncorrelated.
To do this, we have written a helper function, `plot_icar_corr_matrix`,
which plots just the upper half of the correlation matrix of `phi`,
with red indicating positive correlation and blue indicating negative correlation.
The census tracts are listed roughly in geographic order, from north-to-south and
east-to-west, thus adjacent elements along the diagonal are spatially adjacent.

**Python**
```{python}
phi_multicomp = bym2_multicomp_fit.stan_variable("phi")
corr_plot = plot_icar_corr_matrix(phi_multicomp, 'BYM2 spatial correlation', (10, 10))
corr_plot
```

**R**
```{r}
phi_multicomp <- as_draws_matrix(bym2_multicomp_fit$draws("phi"))
plot_icar_corr_matrix(phi_multicomp, 'BYM2 spatial correlation')
```

Plotting the entire correlation matrix puts all covariances on the same scale.
Plotting the component-by-component correlations more clearly shows the spatial structure.
The first three components correspond to Brooklyn and Queens (minus the Rockaways),
Manhattan (minus Roosevelt Island), and the Bronx (minus City Island), respectively.

**Python**
```{python}
corr_plots = []
start_idx = 1
for i in range(N_components):
    end_idx = component_sizes[i] + start_idx
    phi_slice = phi_multicomp[ : , start_idx : end_idx ]
    corr_plots.append(plot_icar_corr_matrix(phi_slice,
        "Spatial Correlation vector phi_"+str(i)+"\nBYM2 multicomp",
	size=(10,10)))
```

```{python}
# Brooklyn and Queens (mainland)
corr_plots[0]
```    
```{python}
# Manhattan
corr_plots[1]
```    
```{python}
# Bronx
corr_plots[2]
```    

**R**
```{r}
corr_plots <- list()
start_idx <- 1
for (i in seq_len(N_components)) {
    end_idx <- start_idx + component_sizes[i] - 1
    phi_slice <- phi_multicomp[, start_idx:end_idx]
    corr_plots[[i]] <- plot_icar_corr_matrix(phi_slice, 
        paste("Spatial Correlation vector phi_", i, "\nBYM2 multicomp", sep = ""))
    start_idx <- end_idx + 1
}
```

```{r}
# Brooklyn and Queens (mainland)
corr_plots[0]
```    
```{r}
# Manhattan
corr_plots[1]
```    
```{r}
# Bronx
corr_plots[2]
```    

### Posterior Predictive Checks

**Python**
```{python}
y_rep_multicomp = bym2_multicomp_fit.stan_variable("y_rep")
print("PPC BYM2\n", ppc_central_interval(y_rep_multicomp, bym2_data['y']))

ppc_plot = ppc_y_yrep_overlay(y_rep_multicomp, bym2_data['y'],
                                'BYM2 multicomp model PPC\ny (blue dot) vs. y_rep (orange 50% central interval, grey full extent)')
ppc_plot
```

**R**
```{r}
y_rep_multicomp <- as_draws_matrix(bym2_multicomp_fit$draws("y_rep"))
cat("PPC BYM2_multicomp")
ppc_central_interval(y_rep_multicomp, bym2_data$y)

ppc_y_yrep_overlay(y_rep_multicomp, bym2_data$y,
                     'BYM2 multicomp model PPC\ny (blue dot) vs. y_rep (orange 50% central interval, grey full extent)')
```

## Model Comparison

With the BYM2_multicomp model, we can use all of the NYC dataset to estimate the regression co-efficients of interest *and* account for the amount of spatial variance and identify strongly correlated regions.
We compare this fit with the Poisson + RE model, which also accounts for the overdispersed nature of the data.

**Python**
```{python}
pois_re_mod = CmdStanModel(stan_file=os.path.join('stan', 'poisson_re.stan'))
pois_re_fit = pois_re_mod.sample(data=bym2_data)
pois_re_summary = pois_re_fit.summary()
pois_re_summary.round(2).loc[
  ['beta_intercept', 'beta0', 'betas[1]', 'betas[2]', 'betas[3]', 'betas[4]', 'sigma']]
```

**R**
```{r}
pois_re_mod = cmdstan_model(stan_file=file.path('stan', 'poisson_re.stan')) 
pois_re_fit = pois_re_mod$sample(data=bym2_data, parallel_chains=cores, refresh=0)
pois_re_fit$print(variables = c('beta_intercept', 'beta0', 'betas[1]', 'betas[2]', 'betas[3]', 'betas[4]', 'sigma'))
```

### Poisson + RE Model PPC coverage, plots

**Python**
```{python}
y_rep_re = pois_re_fit.stan_variable("y_rep")
print("PPC Poisson + RE\n", ppc_central_interval(y_rep_re, bym2_data['y']))

ppc_plot = ppc_y_yrep_overlay(y_rep_re, bym2_data['y'],
                                'Poisson + RE model PPC\ny (blue dot) vs. y_rep (orange 50% central interval, grey full extent)')
ppc_plot
```

**R**
```{r}
y_rep_re <- as_draws_matrix(pois_re_fit$draws("y_rep"))
cat("PPC Poisson + RE")
ppc_central_interval(y_rep_re, bym2_data$y)

ppc_y_yrep_overlay(y_rep_re, bym2_data$y,
                     'Poisson + RE model PPC\ny (blue dot) vs. y_rep (orange 50% central interval, grey full extent)')
```

## Discussion

The BYM2_multicomp model requires a few more kinds of input data to identify the number and size
of the disconnected components and singletons.
It also requires that the data be organized by component, in order to allow for efficient computation.
This, in turn, increases the number of steps required to assemble the data inputs.

Assembling the spatial effects vector `phi` from a series of ICAR components, each of which
has a sum-to-zero constraint for identifiability, requires a by-hand implementation of the
constraining transform for the sum-to-zero vector; this increases the size and complexity
of the model.

However, the alternative is creating a spatial map which doesn't reflect the true
spatial adjacency structure of the data.
This was done for the 2019 analysis of Morris et al.
The results of the regression are reported in Table 4:

## Parameter Estimates from BYM2 Model 

| Parameter       | Mean  | SE Mean | SD   | 2.5%  | 97.5% | N_eff | R-hat |
|--------------- |------:|--------:|-----:|------:|------:|------:|------:|
| Intercept      | -3.5  | 0.0     | 0.5  | -4.5  | -2.5  | 1255  | 1.0   |
| Commute        |  0.5  | 0.0     | 0.2  |  0.2  |  0.9  |  777  | 1.0   |
| Log Income     | -0.1  | 0.0     | 0.0  | -0.2  | -0.0  | 1204  | 1.0   |
| Std Frag Index |  0.1  | 0.0     | 0.0  |  0.0  |  0.1  | 1527  | 1.0   |
| Log Traffic    |  0.0  | 0.0     | 0.0  | -0.0  |  0.0  | 2551  | 1.0   |
| Rho            |  0.4  | 0.0     | 0.1  |  0.3  |  0.5  |  219  | 1.0   |
| Sigma          |  0.8  | 0.0     | 0.0  |  0.8  |  0.9  |  301  | 1.0   |



>>> Because the commute data is recorded as a percentage between 0 and 1
and not on the scale 0 to 100, it is necessary to divide
the "pct_commute" regression coefficient by 100 in order to properly
interpret its contribution.
Thus for every percentage point increase in population commuting by means
other than a private vehicle, there was a exp(0.005) = 0.5% increase in
the expected count of youth pedestrian injuries, controlling for
income, vehicular traffic, social fragmentation, and population.  The
credible interval ranged from a exp(0.002) = 0.2% to exp(0.009) = 0.9%
increase in pedestrian injuries per percentage point increase in
on-foot commuters.  There was a 1.2% decrease in youth pedestrian
injuries per 10% increase in median household income.  Social
fragmentation was also significantly associated with youth pedestrian
injuries, with an exp(0.1) = 10% increase in youth pedestrian injuries
per standard deviation increase in the combined index (i.e. vacancy,
non-owner occupied housing, recent moves, and householder living
alone), controlling for other model covariates.  The credible interval
for the effect of daily traffic included zero in our fully adjusted
model after controlling for social fragmentation and
pedestrian/bicyclist/public transit commute rates.
The parameter sigma, the overall variance of the combined
random effects term was exp(0.8) = 2.2,  indicating
substantial overall variance.
Nearly half of that variance, parameter rho,
was spatially structured exp(0.4) = 49%.

How does this compare with the estimates here?

**Python**
```{python}
bym2_multicomp_summary.round(2).loc[
  ['beta_intercept', 'betas[1]', 'betas[2]', 'betas[3]', 'betas[4]', 'sigma', 'rho']]
```

**R**
```{r}
bym2_multicomp_fit$print(variables = c('beta_intercept', 'betas[1]', 'betas[2]', 'betas[3]', 'betas[4]', 'sigma', 'rho')) 
```

