---
title: "Spatial Data Analysis in Python and R"
format:
  html:
    css: theming/quarto_styles.css 
    syntax-definitions:
      - theming/stan.xml
    toc: true
    toc-location: left
execute:
  eval: false
  keep-ipynb: true
---


## Example Dataset:  NYC traffic accidents, localized to US Census blocks

The dataset we're using is that used in the analysis published in 2019
[Bayesian Hierarchical Spatial Models: Implementing the Besag York Mollié Model in Stan](https://www.sciencedirect.com/science/article/pii/S1877584518301175).

The data consists of motor vehicle collisions in New York City,
as recorded by the NYC Department of Transportation, between the years 2005-2014,
restricted to collisions involving school age children 5-18 years of age as pedestrians.

Each crash was localized to the US Census tract in which it occurred, using boundaries from the 2010 United States Census,
using the [2010 Census block map for New York City](https://data.cityofnewyork.us/City-Government/2010-Census-Blocks/v2h8-6mxf)
The US Census tract is the lowest level of Census data resolution.
A tract contains between 2,500 and 8,000 residents and boundaries that follow visible features.
According to the [Census Bureau](https://www2.census.gov/geo/pdfs/reference/GARM/Ch10GARM.pdf)

> When first established, census tracts are to be as homogeneous as possible with respect to population characteristics, economic status, and living condition

Study Data, per tract:

* collision count
* youth age population
* Census Bureau's American Community Survey (ACS) data for 5-year estimates of:
  + median household income
  + the percentage of commuters who traveled to work by means other than a private vehicle (i.e., by walking, bicycling, or using public transportation)
  + an index of social fragmentation based on ACS estimates of vacant housing units, householders living alone, non-owner occupied housing units, and population having moved within the previous year.
  + street level annual average daily traffic (AADT) data from the New York State Department of Transportation (assigning each census tract the maximum AADT value over the streets in that tract).
* US Census data per tract, including areal geometry (bounding polygons)


## Spatial Data manipulation in R and Python

Geographic information systems (GIS) data is any item which has a geographic location, either a single point or a set of bounding polygons.
Cartographic data (maps) are encoded as a set of records, one per map region.  The [shapefile format](https://en.wikipedia.org/wiki/Shapefile) is an open specification used to insure interoperatility among GIS software packages.  When items in a dataset contain location labels, it is necessary to obtain a set of shapefiles for the corresponding map.
To compute with GIS data, we need packages which can do the geographic math.

- [GeoPandas](https://geopandas.org/en/stable/) extends the datatypes used by pandas to allow spatial operations on geometric types,
which are equivalent to the operations provided by R's [`sf`](https://r-spatial.github.io/sf/) package.
- [`libpysal`](https://pysal.org/libpysal/) is spatial analysis package which can analyze distance between locations; operations equivalent to R's [`spdep`](https://r-spatial.github.io/spdep/) package.

To visualize GIS data, we need to be able to draw maps and spatial graphs.

- In R, ggplot2 can produce beautiful maps for GIS data.
- The python equivalent of R's `ggplot2` is [`plotnine`](https://plotnine.org/).
- The [`splot`](https://pysal.org/notebooks/viz/splot/intro.html) provides additional plotting functions for spatial graph networks.


## Preliminary Data Analysis


### Notebook Startup: Load Packages / Libraries

**Python**
```{python}
# import all libraries used in this notebook
import os
import numpy as np
import pandas as pd
import geopandas as gpd
import libpysal as sa
import matplotlib as plt
import splot as splt
import plotnine as p9
%matplotlib inline


# suppress plotnine warnings
import warnings
warnings.filterwarnings('ignore')

# setup plotnine look and feel global
p9.theme_set(
  p9.theme_grey() + 
  p9.theme(text=p9.element_text(size=10),
        plot_title=p9.element_text(size=14),
        axis_title_x=p9.element_text(size=12),
        axis_title_y=p9.element_text(size=12),
        axis_text_x=p9.element_text(size=8),
        axis_text_y=p9.element_text(size=8)
       )
)

# useful themes
xlabels_90 = p9.theme(axis_text_x = p9.element_text(angle=90, hjust=1))
map_theme =  p9.theme(figure_size=(7,6),
                 axis_text_x=p9.element_blank(),
                 axis_ticks_x=p9.element_blank(),
                 axis_text_y=p9.element_blank(),
                 axis_ticks_y=p9.element_blank())
```

**R**
```{r}
library(sf)
library(spdep) |> suppressPackageStartupMessages()
library(ggplot2)
library(tidyverse) |> suppressPackageStartupMessages()
```



### Load study data

The study data and GIS data have been assembled into a single GeoJSON file.
Let's load the dataset and run a few preliminary visualizations.

**Python**
```{python}
nyc_geodata = gpd.read_file(os.path.join('data', 'nyc_study.geojson'))
print(type(nyc_geodata))
print(nyc_geodata.columns)
print(nyc_geodata['BoroName'].value_counts())
```

**R**
```{r}
nyc_geodata = st_read("data/nyc_study.geojson")
class(nyc_geodata)
names(nyc_geodata)
table(nyc_geodata$BoroName)
```

### Plot population per tract, accidents per tract.

**Python**
```{python}
(
    p9.ggplot(nyc_geodata, mapping=p9.aes(x='kids_per_sq_mi', y='count')) +
    p9.geom_point() +
    p9.geom_jitter(width=0.001, alpha=0.2) +
    p9.geom_smooth(method="lm", color="blue")
)
```

**R**
```{r}
ggplot(nyc_geodata, aes(x = kids_per_sq_mi, y = count)) +
  geom_point() + 
  geom_jitter(width = 0.001, alpha = 0.2) + 
  geom_smooth(method = "lm", color = "blue")
```

### Visualize by Map Areas

#### Traffic accidents - kids hit by cars

**Python**
```{python}
(
    p9.ggplot(nyc_geodata) +
    p9.geom_map(p9.aes(fill='count')) +
    p9.scale_fill_gradient(low="lightblue", high="red") +
    p9.labs(title="NYC School-Age Traffic Accident Victims by Census Tract") +
    map_theme
)
```

**R**
```{r}
ggplot(data = nyc_geodata) +
geom_sf(aes(fill = count)) +
scale_fill_gradient(low = "lightblue", high = "red") +
labs(title = "NYC School-Age Traffic Accident Victims by Census Tract")
```

#### Population: raw counts, density per square mile

**Python**
```{python}
plot_pop = (
    p9.ggplot(nyc_geodata) +
    p9.geom_map(p9.aes(fill='kid_pop')) +
    p9.scale_fill_gradient(low="lightblue", high="red") +
    p9.labs(title = "NYC Population Ages 5-18 per Census Tract") +
    map_theme
)

plot_pop_density = (
    p9.ggplot(nyc_geodata) +
    p9.geom_map(p9.aes(fill='kids_per_sq_mi')) +
    p9.scale_fill_gradient(low="lightblue", high="red") +
    p9.labs(title = "NYC Population Density Ages 5-18 per Census Tract",
         subtitle = "Density measured as population per square mile") +
    map_theme	 
)

plot_pop.show()
plot_pop_density.show()
```

**R**
```{r}
ggplot(data = nyc_geodata) +
geom_sf(aes(fill = kid_pop)) +
scale_fill_gradient(low = "lightblue", high = "red") +
    labs(title = "NYC Population Ages 5-18 Density by Census Tract",
         subtitle = "Density measured as population per square mile")


ggplot(data = nyc_geodata) +
geom_sf(aes(fill = kids_per_sq_mi)) +
scale_fill_gradient(low = "lightblue", high = "red") +
    labs(title = "NYC Population Ages 5-18 Density by Census Tract",
         subtitle = "Density measured as population per square mile")
```


##  Exercises

Visualize `med_hh_inc` (median household income), `frag_index` (fragmentation index), and `traffic`.
